---
title: "Swim_Times_EDA"
output: html_document
date: "2023-03-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r}
# put years into one column 
getwd()
setwd("C:\\Users\\Default\\Desktop\\BruinSportsAnalytics\\")
df_50 <- read.csv("Olympedia - Mens 50 Free.csv", header = TRUE)
df_100 <- read.csv("Olympedia - Mens 100 Free.csv", header = TRUE)
df_200 <- read.csv("Olympedia - Mens 200m  Free .csv", header = TRUE) # ok
df_200$Event <- 200
df_400 <- read.csv("Final_Olympedia - Mens 400m  Free .csv", header = TRUE) # fill NA not working?
df_1500 <- read.csv("Final_Olympedia - Mens 1500m Free.csv", header = TRUE)


#df_100 <- read.csv("C:\\Users\\Desktop\\Downloads\\BruinSportsAnalytics\\Olympedia - Mens 100 Free.csv", header = TRUE)
#df_200 <- read.csv("C:\\Users\\Desktop\\Downloads\\Olympedia - Mens 200m  Free .csv", header = TRUE) # ok
#df_400 <- read.csv("C:\\Users\\Desktop\\Downloads\\BruinSportsAnalytics\\Olympedia - Mens 400m Free.csv", header = TRUE)

#"C:\Users\Default\Desktop\BruinSportsAnalytics\Olympedia - Mens 50 Free.csv"
#"C:\Users\Desktop\Downloads\Olympedia - Mens 200m  Free .csv"

df <- rbind(df_50, df_100, df_200, df_400, df_1500)
df_50_and_100 <- rbind(df_50, df_100)
df_200_400_1500 <- rbind(df_200, df_400, df_1500)

df_50_and_100[df_50_and_100 == ""] <- NA
df_200_400_1500[df_200_400_1500 == ""] <- NA
#tail(df_50_and_100)


# library(chron)

split_R1_50 <- strsplit(df_50_and_100$R1, split = " ")
df_50_and_100$R1 <- as.numeric(sapply(split_R1_50, function(x){
  x[1]
})) # select for the first element in each list component using substr OR convert to dataframe?

names(df_50_and_100)[5] <- "R1 (sec)" 
tail(df_50_and_100)

head(df_200_400_1500$R1)
split_R1_200 <- strsplit(df_200_400_1500$R1, split = "[: ]")
head(split_R1_200)
df_200_400_1500$R1 <- sapply(split_R1_200, function(x){ # character valus
  as.numeric(x[2]) / 60 + as.numeric(x[1]) # ; have to make sure everything is numeric
}) # select for the first element in each list component using substr OR convert to dataframe?

names(df_200_400_1500)[5] <- "R1 (min)" 

df_200_400_1500$RT <- as.numeric(df_200_400_1500$RT)
tail(df_200_400_1500)

# Make sure Event column is a category or factor variable
df_200_400_1500$Event <- factor(df_200_400_1500$Event)

#df_200_400_1500 <- df_200_400_1500[complete.cases(df_200_400_1500[, c("R1 (min)", "RT", "Age", "Height", "Weight")]), ]
# df_200_400_1500[df_200_400_1500$`R1 (min)`!= NA | df_200_400_1500$RT != NA,]
#install.packages('tidyr')

library("tidyr")

# remove rows that have at least one NA
df_200_400_1500 <- df_200_400_1500 %>% drop_na() 
tail(df_200_400_1500) #360

df_50_and_100 <- df_50_and_100 %>% drop_na()
tail(df_50_and_100) #385

# hey so i noticed that some obs has NA for one of the factors (RT, Age, Height, Weight). theres some that has RT missing only and age, height, weight are intact (100 out 416 rows). ..can I keep those rows or will that skew the line plot?

# follow up with what if there's obs that has missing height and weight. 

# interviews


#tail(df_50_and_100) #420 observations

# exporting dataframe to csv
write.csv(df_200_and_400,"C:\\Users\\Default\\Desktop\\BruinSportsAnalytics\\200_400_df.csv", row.names=FALSE)
write.csv(df_50_and_100, "C:\\Users\\Default\\Desktop\\BruinSportsAnalytics\\50_100_df.csv", row.names=FALSE)

# 48.64 = 48.64 seconds
#tidyverse filter, select, arrange, group_by, summarize, court

```
```{r}
# compute the bmi for each metric, but wouldn't they all be around the same since they are athletes?
# compute the average bmi of all athletes and use that as comparison
# sapply to find average height and weight


tail(df_200_400_1500)
summary(df_200_400_1500)


```

```{r}
# DON'T INCLUDE IN ARTICLE
# color points by events
# make a bar graph, but have to split up age into intervals
df_list <- split(df_200_400_1500, df_200_400_1500$Event)
df_list$`200`

par(mfrow = c(2, 3))

hist(df_list$`200`$Age, prob = TRUE)
lines(density(df_list$`200`$Age), lwd = 2, col = "blue")

hist(df_list$`200`$Height, prob = TRUE)
lines(density(df_list$`200`$Height), lwd = 2, col = "blue")

hist(df_list$`200`$Weight, prob = TRUE)
lines(density(df_list$`200`$Weight), lwd = 2, col = "blue")

hist(df_list$`200`$RT, prob = TRUE)
lines(density(df_list$`200`$RT), lwd = 2, col = "blue")

hist(df_list$`200`$bmi, prob = TRUE)
lines(density(df_list$`200`$bmi), lwd = 2, col = "blue")

# Conclusion: use spearman's just to be safe since it works for linear correlation
```
```{r}
# DON'T INCLUDE IN ARTICLE
# color points by events
# make a bar graph, but have to split up age into intervals

par(mfrow = c(2, 3))

hist(df_list$`1500`$Age, prob = TRUE)
lines(density(df_list$`1500`$Age), lwd = 2, col = "blue")

hist(df_list$`1500`$Height, prob = TRUE)
lines(density(df_list$`1500`$Height), lwd = 2, col = "blue")

hist(df_list$`1500`$Weight, prob = TRUE)
lines(density(df_list$`1500`$Weight), lwd = 2, col = "blue")

hist(df_list$`1500`$RT, prob = TRUE)
lines(density(df_list$`1500`$RT), lwd = 2, col = "blue")

hist(df_list$`1500`$bmi, prob = TRUE)
lines(density(df_list$`1500`$bmi), lwd = 2, col = "blue")

# Conclusion: use spearman's just to be safe since it works for linear correlation
```


```{r}
# distribution of age is not normal and scatterplot is not showing linear relationship, and treating age as categorical so don't use pearson's or try to fit a linear regression model
model_200 <- lm(`R1 (min)`~Age, data = df_list$`200`)
correlation <- cor(df_list$`200`$Age, df_list$`200`$`R1 (min)`, method = "spearman")
correlation

# pearson's correlation can only be used for normal distribution
# treating age as a categorical variable, notice that data is not normally distributed so use spearman or kendall's correlation
plot(`R1 (min)`~Age, data = df_list$`200`, cex = 0.7, col = "#00AFBB", ylab = "Time (min)", main = "200 m Freestyle") # treating age as a categorical variable, so use spearman or kendall's correlation
text(30, 2, labels = paste("r = ", round(correlation, 2)), cex = 0.7, col = "red")
lines(loess.smooth(df_list$`200`$Age, df_list$`200`$`R1 (min)`, col="red")) #

model_1500 <- lm(`R1 (min)`~Age, data = df_list$`1500`)
correlation2 <- cor(df_list$`1500`$Age, df_list$`1500`$`R1 (min)`, method = "spearman")
correlation2

# pearson's correlation can only be used for normal distribution
# treating age as a categorical variable, notice that data is not normally distributed so use spearman or kendall's correlation
plot(`R1 (min)`~Age, data = df_list$`1500`, cex = 0.7, col = "#00AFBB", ylab = "Time (min)", main = "1500 m Freestyle") # treating age as a categorical variable, so use spearman or kendall's correlation
text(30, 2, labels = paste("r = ", round(correlation2, 2)), cex = 0.7, col = "red")
lines(loess.smooth(df_list$`1500`$Age, df_list$`1500`$`R1 (min)`, col="red")) #


```



```{r, eval = FALSE}
#REMOVE THIS PART
# make a bar graph or cluster plot
# make sure that each age range contains roughly the same number of observations
# bar graph only shows the average 

num_ranges <- 3
cut_points <- quantile(df_list$`200`$Age, probs = seq(0, 1, length.out = num_ranges + 1), na.rm = TRUE)
age_ranges <- cut(df_list$`200`$Age, breaks = cut_points, include.lowest = TRUE)
ranges <- levels(age_ranges)

df_list$`200`$Age_Range <- age_ranges # store to a new column called Age Range
tail(df_list$`200`)

# ranges: [16,21] (21,24] (24,32]

# make a side by side box plot but doesnt make sense to find median since time is diff --> bar plot, find the mean
# to show the mean of each category
install.packages("ggplot2")
library(ggplot2)
print(tapply(df_list$`200`$`R1 (min)`, df_list$`200`$Age_Range, mean))

ggplot(df_list$`200`, aes(x = df_list$`200`$Age_Range, y = df_list$`200`$`R1 (min)`)) + 
  stat_summary(fun = mean, geom = "bar", color = "black") +
  ggtitle("200 m Freestyle") +
  xlab("Age Ranges") +
  ylab("Average Time (min)") +
  scale_x_discrete(labels = c("16-21", "21-24", "24-32")) 
  
# mean R1 time for each age range

# rearrange the df by age ranges

```



```{r}
# when to use mean vs median: mean for normal distribution and no outliers


num_ranges <- 3
cut_points <- quantile(df_list$`200`$Age, probs = seq(0, 1, length.out = num_ranges + 1), na.rm = TRUE)
age_ranges <- cut(df_list$`200`$Age, breaks = cut_points, include.lowest = TRUE)
ranges <- levels(age_ranges)
ranges

df_list$`200`$Age_Range <- age_ranges # store to a new column called Age Range
tail(df_list$`200`)

ggplot(df_list$`200`, aes(x=df_list$`200`$Age_Range, y = df_list$`200`$`R1 (min)`)) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = "point", shape = 18, size = 2.5, color = "#E7B800") +
  ggtitle("200 m Freestyle") +
  xlab("Age") +
  ylab("Time (min)") +
  scale_x_discrete(labels = c("16-21", "22-24", "25-32"))

table(age_ranges)

## test




```

```{r}
# comparing correlation coefficient across different distances

#1500 m scatterplot and box plots
# make the same ranges for 1500 m event, will most likely have different ranges for 
model_1500 <- lm(`R1 (min)`~Age, data = df_list$`1500`)
correlation_1500 <- cor(df_list$`1500`$Age, df_list$`1500`$`R1 (min)`, method = "spearman")
correlation_1500

install.packages("minerva")
library(minerva)
mic_1500 <- mine(x=df_list$`1500`$Age, y=df_list$`1500`$`R1 (min)`)
mic_1500 <- mic_1500$MIC

mic_200 <- mine(x=df_list$`200`$Age, y=df_list$`200`$`R1 (min)`)
mic_200 <- mic_200$MIC

mic_200 <- mine(x=df_list$`200`$Age, y=df_list$`200`$`R1 (min)`)
mic_200 <- mic_200$MIC

mic_400 <- mine(x=df_list$`400`$Age, y=df_list$`400`$`R1 (min)`)
mic_400 <- mic_400$MIC

```

```{r}
# pearson's correlation can only be used for normal distribution
# treating age as a categorical variable, notice that data is not normally distributed so use spearman or kendall's correlation
plot(`R1 (min)`~Age, data = df_list$`1500`, cex = 0.7, col = "#FC4E07", ylab = "Time (min)", main = "1500 m Freestyle") # treating age as a categorical variable, so use spearman or kendall's correlatiojn
text(30, 16, labels = paste("r = ", round(correlation, 2)), cex = 0.7, col = "black")
#lines(loess.smooth(df_list$`1500`$Age, df_list$`1500`$`R1 (min)`, col="black")) #

# scatter_1500 <- ggplot(df_list$`1500`, aes(x=df_list$`1500`$Age, y=df_list$`1500`$`R1 (min)`)) +
#  geom_point() +
#  geom_smooth(method = "spearman") +
#  geom_text(aes(label = label), hjust = 0, vjust = 0)


# ranges <- levels(age_ranges)

ranges_1500 <- cut(df_list$`1500`$Age, breaks = c(13, 19, 24, 32)) # 19, 21, 24
# table(ranges_1500) # use this cutoff --use these 4 cutoffs
df_list$`1500`$Age_Range <- ranges_1500

ggplot(df_list$`1500`, aes(x=df_list$`1500`$Age_Range, y = df_list$`1500`$`R1 (min)`)) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = "point", shape = 18, size = 2.5, color = "#E7B800") +
  ggtitle("1500 m Freestyle") +
  xlab("Age") +
  ylab("Time (min)") +
  scale_x_discrete(labels = c("14-19", "20-24", "25-32")) 

```

```{r}
# compute mic (maximal information coefficient)
```

```{r}
# make a function for age function
#REMOVE STUFF FROM BELOW
graph_Age <- function(df, ggtitle, labels) {
  model_200 <- lm(`R1 (min)`~Age, data = df_list$`200`)
  correlation <- cor(df_list$`200`$Age, df_list$`200`$`R1 (min)`, method = "spearman")
  
  # pearson's correlation can only be used for normal distribution
  # treating age as a categorical variable, notice that data is not normally distributed so use spearman or kendall's correlation
  plot(`R1 (min)`~Age, data = df_list$`200`, cex = 0.7, col = "#00AFBB", ylab = "Time (min)", main = "200 m Freestyle") # treating age as a categorical variable, so use spearman or kendall's correlation
  text(30, 2, labels = paste("r = ", round(correlation, 2)), cex = 0.7, col = "red")
  lines(loess.smooth(df_list$`200`$Age, df_list$`200`$`R1 (min)`, col="red")) #
  
  num_ranges <- 3
  cut_points <- quantile(df_list$`200`$Age, probs = seq(0, 1, length.out = num_ranges + 1), na.rm = TRUE)
  age_ranges <- cut(df_list$`200`$Age, breaks = cut_points, include.lowest = TRUE)
  ranges <- levels(age_ranges)
  
  df_list$`200`$Age_Range <- age_ranges # store to a new column called Age Range
  tail(df_list$`200`)
  
  print(tapply(df_list$`200`$`R1 (min)`, df_list$`200`$Age_Range, mean))
  
  ggplot(df_list$`200`, aes(x = df_list$`200`$Age_Range, y = df_list$`200`$`R1 (min)`)) + 
    stat_summary(fun = mean, geom = "bar", color = "black") +
    ggtitle("200 m Freestyle") +
    xlab("Age Ranges") +
    ylab("Average Time (min)") +
    scale_x_discrete(labels = c("16-21", "21-24", "24-32")) 
}

# REMOVE THIS BELOW

model_200 <- lm(`R1 (min)`~Age, data = df_list$`200`)
correlation <- cor(df_list$`200`$Age, df_list$`200`$`R1 (min)`, method = "spearman")

# pearson's correlation can only be used for normal distribution
# treating age as a categorical variable, notice that data is not normally distributed so use spearman or kendall's correlation
plot(`R1 (min)`~Age, data = df_list$`200`, cex = 0.7, col = "#00AFBB", ylab = "Time (min)", main = "200 m Freestyle") # treating age as a categorical variable, so use spearman or kendall's correlation
text(30, 2, labels = paste("r = ", round(correlation, 2)), cex = 0.7, col = "red")
lines(loess.smooth(df_list$`200`$Age, df_list$`200`$`R1 (min)`, col="red")) #

print(tapply(df_list$`200`$`R1 (min)`, df_list$`200`$Age_Range, mean))

ggplot(df_list$`200`, aes(x = df_list$`200`$Age_Range, y = df_list$`200`$`R1 (min)`)) + 
  stat_summary(fun = mean, geom = "bar", color = "black") +
  ggtitle("200 m Freestyle") +
  xlab("Age Ranges") +
  ylab("Average Time (min)") +
  scale_x_discrete(labels = c("16-21", "21-24", "24-32")) 
```

```{r}
# color points by events

my_cols <- c("#00AFBB", "#E7B800", "#FC4E07")
pairs(df_200_400_1500[c("R1 (min)", "RT", "Age", "Height", "Weight", "bmi")], col = my_cols[df_200_400_1500$Event], pch = 16, cex = 0.5)

#first row: factors vs R1-- seems like there is no relationship


```
The first row of the scatterplot matrix represents the R1 (min) plotted against all the variables (RT, Age, Height, Weight, and bmi) which is what we're interested in looking at. Blue, yellow, and orange points represent the athletes competing in the 200, 400, and 1500 m events respectively. Looking at the first row, the R1 times vary significantly across the events, scaling the plots too much in the y direction and making it hard for us to see if there's a correlation. Our next steps would be to make line plots for each event so that we can see the correlation better:

```{r}
# scatterplot matrix to see how the correlations compare with each other
df_list <- split(df_200_400_1500, df_200_400_1500$Event)
df_list$`200`

pairs(df_list$`200`[c("R1 (min)", "RT", "Age", "Height", "Weight", "bmi")], col = "#00AFBB", pch = 16, cex = 0.5)
```


```{r}
#50 meters line correlations for RT vs R1

head(df_50_and_100)
df_50 <- df_50_and_100[df_50_and_100$Event == 50, ]
tail(df_50)

model_50 <- lm(R1~RT, data = df_50)
correlation2 <- cor(df_50$RT, df_50$R1, method = "spearman")
correlation2


# pearson's correlation can only be used for normal distribution
# treating age as a categorical variable, notice that data is not normally distributed so use spearman or kendall's correlation
plot(R1~RT, data = df_50, cex = 0.7, col = "#00AFBB", ylab = "Time (sec)", main = "50 m Freestyle") # treating age as a categorical variable, so use spearman or kendall's correlation
text(30, 2, labels = paste("r = ", round(correlation2, 2)), cex = 0.7, col = "red")
lines(loess.smooth(df_50$RT, df_50$R1, col="red")) #

# 200 m line correlations for RT vs R1




```
```{r}

model_200 <- lm(`R1 (min)`~RT, data = df_list$`200`)
correlation200 <- cor(df_list$`200`$RT, df_list$`200`$`R1 (min)`, method = "spearman")
correlation200

plot(`R1 (min)`~RT, data = df_list$`200`, cex = 0.7, col = "#00AFBB", ylab = "Time (min)", main = "200 m Freestyle") # treating age as a categorical variable, so use spearman or kendall's correlation
text(30, 2, labels = paste("r = ", round(correlation200, 2)), cex = 0.7, col = "red")
lines(loess.smooth(df_list$`200`$RT, df_list$`200`$`R1 (min)`, col="red")) #

```
```{r}
model_1500 <- lm(`R1 (min)`~RT, data = df_list$`1500`)
correlation1500 <- cor(df_list$`1500`$RT, df_list$`1500`$`R1 (min)`, method = "spearman")
correlation1500

plot(`R1 (min)`~RT, data = df_list$`1500`, cex = 0.7, col = "#00AFBB", ylab = "Time (min)", xlab = "Reaction Time", main = "1500 m Freestyle") # treating age as a categorical variable, so use spearman or kendall's correlation
text(30, 2, labels = paste("r = ", round(correlation200, 2)), cex = 0.7, col = "red")
lines(loess.smooth(df_list$`1500`$RT, df_list$`1500`$`R1 (min)`, col="red")) #

```

```{r}
pairs(df_list$`400`[c("R1 (min)", "RT", "Age", "Height", "Weight", "bmi")], col = "#FC4E07", pch = 16, cex = 0.5)

# maybe plot linear regression line
```
How do the correlations vary across length of event? Just from looking at the scatterplot matrix, there seems to be stronger correlations between R1 and Age and R1 and height for the 1500 m compared to the 200 m
[bar graph of correlation coefficients, divided by level (the 3 events)]
```{r}


# compute the bmi for each metric, but wouldn't they all be around the same since they are athletes?
# compute the average bmi of all athletes and use that as comparison
# sapply to find average height and weight
sapply(df_200_400_1500[c("R1 (min)", "RT", "Age", "Height", "Weight", "bmi")], mean, simply = FALSE)

mean(df_200_400_1500$bmi) # average bmi of all athletes
sd(df_200_400_1500$bmi) # 1.56404

tail(df_200_400_1500) # maybe split the df by bmi (high bmi compared to the average and low bmi compared to the average)

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
